\section{Introduction}

First we can ask: what is a good proof? Depends on the context. A good proof in an exam is a proof that convinces your teacher. A good proof in a paper is a proof that convinces the reviewer. There is some inherent subjectivity in that. Is there a way that veridity of proofs could be non-subjective?

One idea for this is: can I give a proof to a computer and it tell me if this proof is ok? Yes, but you have to have a way to tell the computer what your proof is.

How can we do this? How can a computer understand math? and proofs? Well, we need a language for this. What is a language we can use to write mathematics so that a computer can understand?

The language we are used to to write mathematics is the Zermelo-Fraenkel axioms with the axiom of choice.

% todo: add something else about ZFC here like a brief description

Is this a good language to explain mathematics to a computer?

% todo:
I know the answer is "maybe not" but im not sure why. Also need to investigate it.
What are the difficulties of implementing ZFC if any?

Why do we choose CIC instead? What is CIC? What are the main differences between the two? 

Lean is a functional programming language and a proof verification assistant. -> based in CIC.

% todo: (maybe this could be a footnote)
How can it be both those things at once? I dont know and need to investigate it.


\subsection{Plan of study}

The objective of these notes is:
\begin{itemize}
  \item That you understand what CIC is.
  \item That you can begin proving some small theorems in Lean, while understanding why Lean allows you to do this. 
\end{itemize}

Chapters 2 through 5 will help you achieve the first point. You may see CIC as built by adding things on top of other things:

(schema of how you build CIC by taking lambda calculus, adding types, then adding dependent types, then adding inductive types)

Chapter 6 will focus on the second point. We will learn what the Curry--Howard isomorphism is and how we can start building proofs.