\section{The untyped lambda calculus}

So in the XX century people became very obsessed with formal mathematics. I mean there always have been some sort of inclination to make math more and more rigurous. In the 1920s, it became clear that the attempts that had been made to clarify the foundations of mathematics suffered from paradoxes an inconsistencies (see Russel's paradox). Thus, German mathematician David Hilbert formulated his program, Hilbert program. In his program he essentially was saying: ok we need to provide secure foundations for all mathematics. This means we need to find some formalism such that all mathematics can be written in a precise formal language, and manipulated according to well defined rules, and we need that all mathematical statements can be proven, no contradiction can be obtained and there should be an algorithm to decide the truth or falsity of any mathematical statement.

This last bit is called the Entscheidungsproblem or decision problem. 

Now several people tried to tackle this problem, in particular we can look at Alan Turing and Alonzo Church. Both mathematician started working on separate versions of their solution for this problem.

Now both of them thought: ok so before we can solve this problem we need to formalize this problem. How can we define an "algorithm"? An algorithm is a procedure that takes an input and produces an output; i.e. a function. But a function can be theoretical, so how can we define a computable function?

To answer this, Alan Turing came up with his Turing machine model, and he says: ok a computable function is one that can be defined by a Turing machine.

On the other hand Church defines a formal system called the lambda calculus and says: a computable function is one that can be defined by a lambda term.

The lambda calculus is particularly interesting for us because, unlike Turing machines, it is a very simple and expressive formal system. I mean: with very few and simple rules you can construct all computable functions. So this is exactly the type of thing we are interested in if we want a computer to understand mathematics.

Also, it is based purely on functions and their application, which is much closer to the kind of mathematical objects we are already familiar with, and easier to integrate with logical systems.

Moreover, the lambda calculus is not only a model of computation, but it also interacts very naturally with logic. In particular, there is a deep connection between lambda terms and proofs, known as the Curry--Howard correspondence, which allows us to interpret types as propositions and terms as proofs. But more on this later.

\subsection{Constructing terms}

So we are interested in constructing lambda-terms. A lambda terms is simply a computable function or you could think of it as a description of an algorithm.

Lambda-terms are described recursively as follows. First we will assume we have an infinite set of variables $\{x_1, x_2, \dots\}$. Then:

\begin{itemize}
  \item (Base case) Any variable is a lambda-term.
  \item (Recursive case) If $x$ is a variable and $t, t'$ are lambda-terms, then
  \begin{itemize}
    \item $\pilambda x. t$ is a lambda-term (abstraction)
    \item $tt'$ is a lambda-term (yuxtaposition or application)
  \end{itemize}
\end{itemize}

Now at first sight these do not look like functions at all. Remember we are also interested in giving rules that mathematicians can use to manipulate mathematical objects. We now have our mathematical object, the lambda-terms, and we want to give formal rules on how to manipulate them.

I will now try to convince you that, by defining the following rules, we are defining a behaviour that resembles our intuitive idea of a function. I.e. I want to convince you that you can understand $\lambda x. t$ as an expression $f : x \mapsto t$ that you are more used to.

Let $x$ be a variable and $t, t'$ be two any valid lambda terms.

\begin{highlight}[Exercise]
Convince yourself that these are valid lambda-terms:

\begin{itemize}
  \item $\pilambda x. t$
  \item $(\pilambda x. t) t'$.
\end{itemize}
\end{highlight}

First rule: alpha conversion

$$
\pilambda x. t =_\alpha \pilambda y. t, 
$$

whenever $y$ is a variable that did not appear in $t$ already.

\begin{highlight}[Important note]
If $y$ appears in $t$, we could have sometihng like this:

$$\pilambda x. y x =_\alpha \pilambda y. y y$$

which is clearly not the same function.
\end{highlight}

Now this reflects the idea that the actual name of the parameters in a function
definition should be irrelevant. I.e. the function $f(x) = x$ is exactly the same function as $f(y) = y$.

Second rule: beta reduction

$$
(\pilambda x. t)t' =_\beta t[x \mapsto t']
$$

\begin{highlight}[Definition]
$
t[x \mapsto t']
$
is defined recursively, but essentially the thing is "take every instance of $x$ in $t$ and substitute it by $t'$. For example:

$$
(x + y)[x \mapsto 3] = 3 + y
$$

The formal definition is given by: let $x \neq y$ variables and $e, t,t'$ terms.
$$
\begin{array}{rcl}
  x[x \mapsto e] &=& e\\
  y[x \mapsto e] &=& y\\
  (\pilambda z. t)[x \mapsto e] &=& \pilambda z. t[x \mapsto e] \\
  (tt')[x \mapsto e] &=& t[x \mapsto e]t'[x \mapsto e]
\end{array}
$$
\end{highlight}

This second rule reflects the idea of function aplication. For example a function $f : x \mapsto f(x)$ applied to the number $3$ is the result of changing every instance of the variable $x$ for $3$ in the definition of $f$.

\subsection{Pros}

Even with an extremely small syntax and rules for rewriting, we can write rich mathematics. This is great because this is exactly the sort of thing a computer can work with, since we have a very clear idea of what are valid mathematical objects (the lambda terms) and to do computations is simply to rewrite; since everything is a function here, computation comes naturally. This is less true for ZFC, we still have rules but computation is not something primitive, it is simulated. (i should investigate more on this last sentence to at least be able to explain).

Some examples of mathematical objects we can encode as functions using the lambda calculus.

Bools can be described as:

$$
\begin{array}{c}
  \operatorname{TRUE} = \pilambda x.\pilambda y. x \\
  \operatorname{FALSE} = \pilambda x.\pilambda y. Y
\end{array}
$$

The key intuition is that a boolean can be understood as a choice between (any) two alternatives.

From this we can enconde conditionals, i.e. we encode if b then x else y as:

$$
  \operatorname{COND} = \pilambda b. \pilambda x. \pilambda y. b x y
$$

Note that conditionals (controw-flow) is a very basic building block of computation.

Natural numbers can be described as:

We understand a natural number as “repeat a function n times”

Note: in ZFC, natural numbers are sets (von Neumann's naturals). So for example adding two numbers is not really our intuitive idea of actual addition. What is the computation we perform? We are simulating addition.

Even example. Addition example.


\subsection{Exercises}

Important but I want to add exercises that might be interesting to understand something about Lean and i dont know how that could look like for now. Should I just add exercises on lambda calculus so as to get used to it?

Exercise 1. Are the following well-formed lambda terms?

Exercise 2. Reduce the following expressions

Exercise 3. Define AND and OR for booleans. Notice that this means that we can encode logic in lambda calculus without having it as an extra resource!!

Exercise 4. Define ODD. Define MULTIPLE\_OF\_THREE. on naturals.

Exercise 5. Define multiplication on naturals. 

Exercise 6. Devise an encoding for pairs in the lambda calculus. Provide lambda terms PAIR and the projections FST and SND.

\subsection{Cons}

At this point, it might seem that the lambda calculus gives us exactly what we were looking for: a very small formal system, with simple rules, that is expressive enough to encode a wide range of mathematical objects and computations.

However, this expressiveness comes at a cost. The system is extremely permissive: essentially any syntactic combination of terms is allowed, and there is no notion of whether an expression is “well-formed” beyond the grammar itself.

In particular, this means that we can form terms that do not behave like meaningful functions at all.

example:

$$
(\pilambda x. xx) (\pilambda x. xx)
$$

is a well formed term (exercise: convince yourself so). Now try to reduce it. You cannot, every time we try to reduce it, we obtain the same expression again. This is unnaceptable in computation. Remember we're trying to get a computer to understand math. A computer will not realize it has already been here before 23984 times and thus should try a different approach. Thus, we cannot have terms that are infinitely reducible.

Notice that programs don't have to necessarily terminate (you can define partial things with care) but proofs / definitional computation must behave well. At the very least, we need to ensure that the expressions we use to represent mathematical objects behave in a controlled way, and do not lead to meaningless or non-terminating computations.

What is happening here is a form of self-reference: the function is applied to itself, and this leads to an infinite loop. This kind of phenomenon is closely related to classical logical paradoxes.

A famous example is Russell's paradox, which arises when considering the set of all sets that do not contain themselves. Asking whether this set contains itself leads to a contradiction.

Both situations stem from the same underlying issue: allowing unrestricted self-reference leads to pathological behaviour.

For set theory, we simply prohibit self-references and call it a day. However how can we solve this problem? In the next chapter we will find the answer huehue.

